Owl Occupancy in Three of El Salvador's Protected Areas from 2003 through 2013

### 2021-02-28 manuscript_working/ms_working_RMD.Rmd
updated organization, intro/discussion

As apex predators, owls are an important bioindicator of ecosystem health, though much of our understanding comes from temperate systems [@White:2013; @Wan:2018; @Buechley:2019]. In the tropics, particularly in rain and cloud forests, owl studies are more difficult than in temperate regions, regardless of potentially denser populations [@Konig:1999], and neotropical owls are less well-studied than their temperate counterparts [Buechley:2019]. Further, the neotropics were identified by @Buechley:2019 as a priority area for owl research, based on an extensive literature review that identified a relative lack of research attention and a high conservation risk for neotropical owls. As a result, we have limited understanding of neotropical owls’ distribution, ecological requirements, population dynamics, and reproductive behavior [@Clark:1978; @Enriquez:2006; @Perez-Leon:2017; @Rangel-Salazar:2017]. This lack of information poses a challenge to proactive conservation, even as neotropical habitat experiences tremendous anthropogenic pressure and rapid, unpredictable responses to climate change [@Corlett:2012]. In general, while our knowledge about the status and trends of neotropical owl populations is limited, we do know that owl abundance and distribution is decreasing in several regions where species have been added to endangered species lists or have become locally extirpated [@Enriquez:2006].

Like many owl species, neotropical owls are difficult to accurately census due to their cryptic nature and nocturnal habits. Vocalizations of crepuscular and cryptic animals such as owls are commonly leveraged in survey methodology [@Konig:1999]. Owl vocalizations can be diagnostic and are often considered more practically attainable than visual identification. Owls vocalize to communicate with the same species, to threaten or provoke other species, and to delimit territory [@Johnsgard:1988].

Vocalization surveys for nocturnal birds of prey fall generally into two categories. Spontaneous or “passive” surveys involve listening for animal calls with no prior stimulus, whereas provocation or “broadcast” surveys are conducted by listening after a period of playing a recording of animal vocalizations [e.g., @Baumgardt:2019].  The inclusion of broadcast calls in auditory surveys can increase the detection probability of owls, although intra- and inter-species interactions can play a role whether owls are more likely or less likely to call in response to a vocalization simulus [@Enriquez:1997; @Borges:2004; @Flesch:2007; @Baumgardt:2019].

@@

El Salvador is located on the western side of the Central American isthmus and is the smallest and most densely populated Central American country  [@CIA:2020].  It is bordered by the Pacific Ocean, Guatemala and Honduras (Fig. 1XXX). Overall, 13.6% of El Salvador's 20,720 km$^2$ of land is forest and 82% is used for agricultural purposes [@CIA:2020].   We conducted surveys in three protected natural areas located at different elevations (high, mid and low elevations) and in different types of forest vegetation (alluvial, deciduous, semi-deciduous, pine-oak and cloud forest) that represent a portion of the country's diverse ecosystems (Fig. 1xxx, Table 1XXX).

El Imposible National Park (NP) was located 119 km southwest of San Salvador (Fig. 1XXX, Table 1XXX).  Its elevation ranged from 250 to 1425 m above sea level (a.s.l.).  It was the largest NP in the country, covering 3792 ha, and its elevation ranged from 250 to 1425 m a.s.l..  The topography of El Imposible was extremely steep and broken, with many cliffs [@Alvarez:2003].  The area contained deciduous and semi-deciduous forest, secondary growth vegetation on the edges, and former pasture areas that were reforested with native species in 1997.  We established one survey route (EI-1) in the western portion of El Imposible in an area of reforestation and at a low elevation and a second survey route (EI-2) in an area of semi-deciduous forest.

@@

Nancuchiname Forest (797 ha) is a Protected Natural Area located 95 km southeast of San Salvador in the coastal alluvial plain along the Lempa River (Fig. 1XXX).  Its elevation ranged from sea level to 12 m a.s.l.  The area contained deciduous and semi-deciduous forest, second growth vegetation, native species reforestation and a dike surrounded by secondary growth.  In Nancuchiname Forest, we established one survey route (N-1) in alluvial forest along the dike and the other route (N-2) in a semi-deciduous alluvial forest.

*Table 1XX. Route locations and habitats. Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador: El Imposible National Park (EI-1 and EI-2), Montecristo National Park (M-1 and M-2) and Nancuchiname Forest (N-1 and N-2). *

@@

We conducted acoustic surveys on foot following the Guidelines for Nocturnaal Owl Monitoring [@Takats:2001], a similar protocol developed for Costa Rica [@Enriquez:2001], and other raptor research guidelines [@Fuller:1987]. This study's survey protocol was an extension of previous research completed in the western half of El Imposible NP [@West:1988].  We created six 2-km long survey routes (transects) that each included 10 permanent survey stations set at 200 m intervals. Two routes were established in each protected area (Table 1xxx).

@@

Mottled Owls, Spectacled Owls, and Ferruginous Pygmy-Owls had 542, 137, and 187 positive detections over the 11 year period, respectively (Table 2XX). No other owl species had enough detections to analyze species-specific occupancy (Table 2XX). Averaged across years, we found that the probability of occupancy was somewhat higher for Mottled Owls than for Ferruginous Pygmy-owls or Spectacled Owls in both routes of El Imposible and in M-2 route of Montecristo (Fig. 1X). Occupancy probabilities were relatively high across all three species for both routes of Nancuchiname (Fig. 2XX). We found that the probability of occupancy for Mottled Owls was relatively consistent across time in both El Imposible routes and in N-2 (Fig. 3XX), but that it seemed to increase in the last approximately 6 years in M-2 and N-1 (Fig. 3XX). The probability of occupancy in M-1 seemed to have decreased through time for Mottled Owls (Fig. 3XX).

For routes EI-1, EI-2, and M-2, the probability of occupancy for Ferruginous Pygmy-owls and Spectacled Owls stayed relatively low across time other than a higher probability of occupancy for Ferruginous Pygmy-owl in EI-1 in the early years of the surveys (Fig. 3XX). The probability of occupancy for Ferruginous Pygmy-owls was relatively high in both routes of Nancuchaname except in 2008 and 2009 (Fig. 3XX), and the probability of occupancy for the Spectacled Owl fluctuated slightly from year to year in the Nancuchaname routes (Fig. 3XX).

@@

When analyzed by year and route, median species richness ($\text{Richness}^*_{t,h}$) varied from 0 to 5 (Fig. 4XXX). At most, the median species richness was one species higher than the number of species detected at any single route and year; however, the 90% credible intervals of richness added up to 4 new undetected species (e.g., in M-2 during 2004 where we detected zero species (Fig. 4XXX).

@@

Including all detected species in one richness model increased the precision of the estimates of the effect of broadcast calls on detection probability (i.e., compare All Species with individual occupancy results in Fig. 5XXX). When incorporating all owl species, broadcasting calls from Mottled Owls, Pacific Screech-owls, Crested Owls, Black-and-white Owls, and Spectacled Owls increased detection probability; whereas broadcasting the Great Horned Owl calls decreased detection (Fig. 5XXX).

The probability of detection when analyzed with single-species occupancy models varied little during the timeframe before broadcast recordings were played (Pre-broadcast); however, the probability of detecting owls after broadcast calls were played varied by call and between Mottled Owls, Ferruginous Pygmy-owls, and Spectacled Owls (Fig. 5XX). The probability of detecting Mottled Owls increased after broadcasting Mottled Owl, Pacific Screech-owl, Crested Owl, Black-and-white Owl, or Spectacled Owl calls. The probability of detecting Ferruginous Pygmy-owls increased after broadcasting Mottled Owl and Pacific Owl calls, and the probability of detecting Spectacled Owls increased after broadcasting Pacific Screech-owl, Black-and-white Owl, and Spectacled Owl calls (Fig. 5XX).

@@

The goal of this project was to gather baseline data on owl abundance, distribution, and population trends in El Salvador. We established survey routes within three protected natural areas in disparate ecological regions (alluvial plain, forested, cloud forests) and attempted to compete two surveys of each route for each year from 2003 through 2013. Our goal of ten years’ worth of repeated surveys was based on similar studies that have previously been conducted on Mexican spotted owls (*Strix occidentalis*) [@USFWS:1995].  However, our surveys were not always completed twice per year due to poor weather, issues with site access, or other complications. We did not survey during significant weather events, as rain and high winds diminish owl vocalizations and the ability for surveyors to detect owl calls [@Takats:2001; @Andersen:2007].   Overall, our surveys yielded sufficient data to analyze occupancy patters for three species of interest; the Mottled Owl, Ferruginous Pygmy-owl, and Spectacled Owl.

The Mottled Owl is the most common and widespread owl in South America [@Vallely:2018] and they have been documented in a wide range of habitat types and elevations in Mexico [@Enriquez-Rocha:1993]. Mottled Owls were the most frequently detected owl during our surveys. A recent study indicated that Mottled Owl populations in El Salvador have persisted despite habitat loss that has restricted the distribution of several other owl species [@Perez-Leon:2017]. In contrast, @Enriquez:2001 stated that Mottled Owl populations in La Selva may have decreased during the past 30 years.  Our results indicated that Mottled Owls are still relatively common at each of our study areas, although there seemed to be a downward trend in occupancy in M-2 that could warrant further monitoring.

Ferruginous Pygmy-owls were the second most-often detected owls in this study. This species demonstrated fairly low occupancy rates in El Imposible routes, other than in the first two years in EI-1. Ferruginous Pygmy-owls tend to prefer open habitats with a lot of edge, such as the flooded igap\'o forests of Amazonia [@Borges:2004] and heterogenous forests in central Argentina [@Campioni:2013], and their dietary composition is dependent on forest density [@Sarasola:2014]. Lower elevation areas of El Imposible were grazed and dotted with houses in the late 1980s [@West:1988], and this mixed agrarian landscape may have provided suitable habitat which has been reduced by subsequent reforestation efforts that began in 1997. It is possible that Ferruginous Pygmy-owls are responding to a decrease in edge habitat within El Salvador, and we recommend further surveys to determine the risk to Ferruginous Pygmy-owls in response to revegetation management.

Changes in habitat availability are not the only challenge facing El Salvador's owls. @Perez-Leon:2017 stated that there are human activities affecting owl populations in El Salvador, including illegal hunting, trapping, persecution, killing, and wildlife trade. They also stated that, of eight species of owls confiscated from local markets in 1995 to 2008, the most frequent were Mottled Owls and Ferruginous Pygmy-owls. If poaching of these species continues, their populations may be at more risk than our results indicated.

Spectacled Owls followed similar occupancy patterns to that of the Ferruginous Pygmy-owls, with highest levels of occupancy in the riparian Nancuchiname Forest. Spectacled Owls can be found in a range of habitats including coffee plantations and urban parks [@Marin-Gomez:2017]. Spectacled Owls tend to prefer lowland habitat, and they were reported to live in elevations below 900 m a.s.l. in Mexico [@Enriquez-Rocha:1993] and in lowland habitats in Ecuador below 1000 m a.s.l. [@Orihuela-Torres:2018]; however, they have been reportedly able to live up to 2600 m a.s.l. [@Marin-Gomez:2017]. We detected Spectacled Owls in Montecristo National Park, transect M-2, which was on average 1755 m a.s.l., which is potentially near their preferred elevation range. Spectacled Owls have been reported to prefer mammalian prey over reptilian or arthropods [@Orihuela-Torres:2018] and preferentially hunt near fallen logs where small mammals often seek refuge [@Esclarski:2014], but we were unable to test for micro-habitat occupancy in our present study. Thus, further refinement over Spectacled Owl habitat selection could be explored in the future to better understand their ecology and distribution in the Neotropics and specifically El Salvador.

Nancuchiname Forest was the richest in diversity of our three protected areas; however, many of the habitat specialist species were only detected in Montecristo NP (e.g., Whiskered Owl, Great Horned Owl, Fulvous Owl, and c.f. Stygian Owl). A Nancuchiname Forest Management Plan only listed one owl, Ferruginous Pygmy-owl, as being present [@Zepeda:1995], but we confirmed the presence of four additional owl species in that NP. Fulvous Owl is one of El Salvador's 18 endemic bird species [@Komar:2002], and it was detected numerous times in the cloud forest of Montecristo NP (M-1). Fulvous Owl has also been recently photographed in Montecristo NP [@Gonzalez:2017].

Stygian Owl, which often inhabit dense cloud forests [@Enriquez-Rocha:1993], were listed as an owl species expected to reside in El Salvador [@Perez-Leon:2017), but has been previously undetected in the country. On 26 March 2002, after listening to recordings and viewing photographs, the Montecristo NP guards stated that they thought c.f. Stygian owl was present in the park (personal communication). While there was only one c.f. Stygian Owl detected on our survey routes, we did also hear a c.f. Stygian Owl between the survey stations on 25 April XXXXyear. We also heard c.f. Stygian Owl vocalizations twice on M-2 between station points on 20 March 2005 and once between stations in 2007. We did not record any of these vocalizations by c.f. Stygian Owls and detections that occurred between stations or away from survey routes were not included in our results. So while we are confident that c.f. Stygian Owls are present in Montecristo NP, we believe that further documentation of the species would be beneficial for confirmation of the species in El Salvador.

During 1979-1981, in the western portion of El Imposible NP, West (1988) heard or observed six owl species, including two that were not detected during this study (Barn Owls and Black-and-white Owls) in that NP. In 1980, Barn Owls nested in a cave of Loma de Paja Mountain (opposite this study's El Imposible routes). An 5 April 2002, Barn Owls were detected in that same location (Vidal Campos Aguirre, personal communication). Vidal Campos Aguirre also stated that Black-and-white Owls were observed in two areas of the Mistepe River Valley in that same timeframe. It is likely, then, that these two owls still occur in El Imposible, though we were not able to detect them. We also observed a Barn Owl nesting near the N-1 route; however, again, we did not include any visual or auditory detections of owls outside of the survey stations. Barn Owls were detected in both Nancuchiname routes, but Black-and-white Owls were not detected in any routes during our surveys.

Distinguishing individuals by non-invasive means, such as vocal traits, is preferable when species are rare, sensitive to handling, elusive, or when other techniques are unfeasible [@Terry:2005]. Overall, the probability of detection was fairly low for all species (median values $\le 0.50$, even though we used locally-recorded owl vocalizations when possible. In Guatemala, @Gerhardt:1989 found that a series of Mottled Owl vocalizations obtained from a non-local ornithological laboratory could not generate a response from Guatemalan owls. We found that the probability of detecting most owls increased after we broadcasted the vocalizations of Mottled Owls, Pacific Screech-owls, Crested Owls, and Spectacled Owls. Conversely, owls were less likely to respond after we played Great Horned Owl calls. Because owls, like many birds of prey, are sensitive to interactions with other owls [@Enriquez:1997; @Baumgardt:2019] and may change their response to broadcast calls depending on their life stage (e.g., nesting versus courtship) [@Flesch:2007], we recommend that surveyors carefully consider the intra- and inter-species interactions that may affect detectability after playing specific bird vocalizations.

Globally, an estimated 75% of the nearly 250 species of owls are associated with dense and undisturbed forests [@Mikkola:2012].  @Perez-Leon:2017 stated that within El Salvador, natural ecosystems had the most diverse owl populations, specifically habitats comprised of cloud, deciduous, riparian and pine-oak forests.  In addition to areas that are preserved in a natural state, shade-grown coffee production may be providing valuable owl habitat. Coffee plantations account for around 7% of El Salvador's forested land due to a prevalence of shade-cover farming practices  [@Silva:2016].  @Silva:2016 also concluded that almost all primary or closed forest is surrounded by coffee farms, which act as a buffer zone against land use change.  Likewise, @Perez-Leon:2017 stated that coffee plantations are the most important agroecosystem for owl communities’ survival in El Salvador.

This research study was critical because of its long-term study of multiple owl species in an rapidly urbanizing and population-dense country with diverse habitats and relatively little previous research. As we continue to identify areas of the world that are owl research priorities based on the lack of current information leveraged against the potential conservation risk, such as in @Buechley:2019, studies such as ours will be important and hopefully increasing in frequency.

@@

![2XX. Probability of occupancy ($\psi_{t,h}$) summarized by route for Ferruginous Pygmy-owl (FEPO), Mottled Owl (MOOW), and Spectacled Owl (SPEO). Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador: El Imposible National Park (EI-1 and EI-2), Montecristo National Park (M-1 and M-2) and Nancuchiname Forest (N-1 and N-2).  Posteriors were summarized with median $\pm$ 90% credible intervals. \label{psi_means}](../output/figures/psi_means-1.png)

![3XX. Probability of occupancy by route and year for Ferruginous Pygmy-owl (FEPO), Mottled Owl (MOOW), and Spectacled Owl (SPEO). Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador: El Imposible National Park (EI-1 and EI-2), Montecristo National Park (M-1 and M-2) and Nancuchiname Forest (N-1 and N-2). Posteriors were summarized with median $\pm$ 90% credible intervals. \label{figure2}](../output/figures/psi_byYr-1.png)

@@

![5XX. The probabiliity of detecting owls during spontaneous surveys (i.e., pre-broadcast) and after broadcast, depending on the species of owl used for broadcast vocalization. These logistic regression coefficients were evaluated from the richness model, which combined all detected species (All Species) and the single-species occupancy models for Ferruginous Pygmy-owls (FEPO), Mottled Owls (MOOW), and Spectacled Owls (SPEO). Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador. Posteriors were summarized with median $\pm$ 90% credible intervals. The pre-broadcase effect size's credible intervals are represented with two grey horizontal lines on each graph \label{figure6}](../output/figures/p_detection-1.png)

>>>>

As apex predators, owls are an important bioindicator of ecosystem health, though much of our understanding comes from temperate systems [@White:2013; @Wan:2018; @Buechley:2019]. The Neotropics were recently identified as a priority area for owl research because the high conservation risk for neotropical owls has not been supported by enough research attention [@Buechley:2019], but in tropical rain and cloud forests, owl studies are much more difficult than in temperate regions, regardless of potentially denser populations [@Konig:1999]. As a result, we have limited understanding of neotropical owls’ distribution, ecological requirements, population dynamics, and reproductive behavior [@Clark:1978; @Enriquez:2006; @Perez-Leon:2017; @Rangel-Salazar:2017]. This lack of information poses a challenge for proactive conservation, even as neotropical habitat experiences tremendous anthropogenic pressure and rapid, unpredictable responses to climate change [@Corlett:2012]. In general, while our knowledge about the status and trends of neotropical owl populations is limited, we do know that owl abundance and distribution is decreasing in several regions where species have been added to endangered species lists or have become locally extirpated [@Enriquez:2006].

El Salvador is located on the western side of the Central American isthmus and is the smallest and most densely populated Central American country [Fig. 1XXX; @CIA:2020]. Overall, 13.6% of El Salvador's 20,720 km$^2$ of land is forested, and 82% is used for agricultural purposes [@CIA:2020]. A common agroecosystem in El Salvador is shade-grown coffee, which supports forested cover on otherwise agricultural land. Approximately 7% of El Salvador's forested land is comprised of shade-grown coffee, which may be providing an important land-use buffer around El Salvador's protected natural areas [@Silva:2016; @Perez-Leon:2017]. Although shade-grown coffee systems are potentially reducing the impacts of land-use and climate change on the forest-dependent owls of El Salvador [e.g., @Perez-Leon:2017], there is a lack of information about the long-term  dynamics of these populations. Concerns about habitat availability are not the only challenge facing El Salvador's owls. @Perez-Leon:2017 stated that there are human activities affecting owl populations in El Salvador, including illegal hunting, trapping, persecution, killing, and wildlife trade.

@@

We conducted surveys in three protected natural areas in El Salvador (El Imposible National Park, Montecristo National Park, and Nancuchiname Forest) located at different elevations (high, mid and low elevations) and in different types of forest vegetation (alluvial, deciduous, semi-deciduous, pine-oak and cloud forest) that represent a portion of the country's diverse ecosystems (Fig. 1xxx, Table 1XXX). El Imposible National Park (NP) was located 119 km southwest of San Salvador (Fig. 1XXX; Table 1XXX).  Its elevation ranged from 250 to 1425 m above sea level (a.s.l.).  It was the largest NP in the country, covering 3792 ha, and its elevation ranged from 250 to 1425 m a.s.l..  The topography of El Imposible was extremely steep and broken, with many cliffs [@Alvarez:2003].  The area contained deciduous and semi-deciduous forest, secondary growth vegetation on the edges, and former pasture areas that were reforested with native species in 1997.  We established one survey route (EI-1) in the western portion of El Imposible in an area of reforestation and at a low elevation and a second survey route (EI-2) in an area of semi-deciduous forest.

@@

Nancuchiname Forest (797 ha) is a Protected Natural Area located 95 km southeast of San Salvador in the coastal alluvial plain along the Lempa River (Fig. 1XXX; Table 1XXX).  Its elevation ranged from sea level to 12 m a.s.l.  The area contained deciduous and semi-deciduous forest, second growth vegetation, native species reforestation and a dike surrounded by secondary growth.  In Nancuchiname Forest, we established one survey route (N-1) in alluvial forest along the dike and the other route (N-2) in a semi-deciduous alluvial forest.

*Table 1XX. Route locations and habitats. Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador: El Imposible National Park (EI-1 and EI-2), Montecristo National Park (M-1 and M-2), and Nancuchiname Forest (N-1 and N-2). *

@@

We conducted acoustic surveys on foot following the Guidelines for Nocturnal Owl Monitoring [@Takats:2001], a similar protocol developed for Costa Rica [@Enriquez:2001], and other raptor research guidelines [@Fuller:1987]. This study's survey protocol was an extension of previous research completed in the western half of El Imposible NP [@West:1988].  We created six 2-km long survey routes (transects) that each included 10 permanent survey stations set at 200 m intervals. Two routes were established in each protected area (Table 1xxx).

@@

Mottled Owls, Spectacled Owls, and Ferruginous Pygmy-Owls had 542, 137, and 187 positive detections over the 11 year period, respectively (Table 2XX). No other owl species had enough detections to analyze species-specific occupancy (Table 2XX). Averaged across years, we found that the probability of occupancy was somewhat higher for Mottled Owls than for Ferruginous Pygmy-Owls or Spectacled Owls in both routes of El Imposible and in M-2 route of Montecristo (Fig. 1X). Occupancy probabilities were relatively high across all three species for both routes of Nancuchiname (Fig. 2XX). We found that the probability of occupancy for Mottled Owls was relatively consistent across time in both El Imposible routes and in N-2 (Fig. 3XX), but that it seemed to increase in the last approximately 6 years in M-2 and N-1 (Fig. 3XX). The probability of occupancy in M-1 seemed to have decreased through time for Mottled Owls (Fig. 3XX).

For routes EI-1, EI-2, and M-2, the probability of occupancy for Ferruginous Pygmy-Owls and Spectacled Owls stayed relatively low across time other than a higher probability of occupancy for Ferruginous Pygmy-Owl in EI-1 in the early years of the surveys (Fig. 3XX). The probability of occupancy for Ferruginous Pygmy-Owls was relatively high in both routes of Nancuchaname except in 2008 and 2009 (Fig. 3XX), and the probability of occupancy for the Spectacled Owl fluctuated slightly from year to year in the Nancuchaname routes (Fig. 3XX).

@@

When analyzed by year and route, median species richness ($\text{Richness}^*_{t,h}$) varied from 0 to 5 (Fig. 4XXX). At most, the median species richness was one species higher than the number of species detected at any single route and year; however, the 90% credible intervals of richness added up to 4 new undetected species (e.g., in M-2 during 2004 where we detected zero species; Fig. 4XXX).

@@

Including all detected species in one richness model increased the precision of the estimates of the effect of broadcast calls on detection probability (i.e., compare All Species with individual occupancy results in Fig. 5XXX). When incorporating all owl species, broadcasting calls from Mottled Owls, Pacific Screech-Owls, Crested Owls, Black-and-white Owls, and Spectacled Owls increased detection probability; whereas broadcasting the Great Horned Owl calls decreased detection (Fig. 5XXX).

The probability of detection when analyzed with single-species occupancy models varied little during the timeframe before broadcast recordings were played (Pre-broadcast); however, the probability of detecting owls after broadcast calls were played varied by call and between Mottled Owls, Ferruginous Pygmy-Owls, and Spectacled Owls (Fig. 5XX). The probability of detecting Mottled Owls increased after broadcasting Mottled Owl, Pacific Screech-Owl, Crested Owl, Black-and-white Owl, or Spectacled Owl calls. The probability of detecting Ferruginous Pygmy-Owls increased after broadcasting Mottled Owl and Pacific Owl calls, and the probability of detecting Spectacled Owls increased after broadcasting Pacific Screech-Owl, Black-and-white Owl, and Spectacled Owl calls (Fig. 5XX).

@@

The goal of this project was to gather baseline data on owl abundance, distribution, and population trends in El Salvador. We established survey routes within three protected natural areas in disparate ecological regions (alluvial plain, forested, cloud forests) and attempted to complete two surveys passes per route each year from 2003 through 2013. Our surveys were not always completed twice per year in each route because of isses with site access or poor weather, which decreases owl vocalizations and detection [@Takats:2001; @Andersen:2007]. Overall, our surveys yielded sufficient data to analyze occupancy patters for three species of interest; the Mottled Owl, Ferruginous Pygmy-Owl, and Spectacled Owl.

Mottled Owls were the most frequently detected owl during our surveys. Mottled Owls are the most common and widespread owl in South America [@Vallely:2018], and they have been documented in a wide range of habitat types and elevations in Mexico [@Enriquez-Rocha:1993]. A recent study indicated that Mottled Owl populations in El Salvador have persisted despite habitat loss that has restricted the distribution of several other owl species [@Perez-Leon:2017]. In contrast, @Enriquez:2001 stated that Mottled Owl populations in La Selva may have decreased during the past 30 years.  Our results indicated that Mottled Owls are still relatively common at each of our study areas, although there seemed to be a downward trend in occupancy in M-2 that could warrant further monitoring.

Ferruginous Pygmy-Owls were the second most-often detected owls in this study. This species demonstrated fairly low occupancy rates in El Imposible routes, other than in the first two years in EI-1. Ferruginous Pygmy-Owls tend to prefer open habitats with a lot of edge, such as the flooded $\text{igap\'o}$ forests of Amazonia [@Borges:2004] and heterogenous forests in central Argentina [@Campioni:2013]. Further, the dietary composition of Ferruginous Pygmy-Owls varies based on forest density [@Sarasola:2014]. Lower elevation areas of El Imposible were grazed and dotted with houses in the late 1980s [@West:1988], and this mixed agrarian landscape may have provided suitable habitat which has been reduced by subsequent reforestation efforts that began in 1997. It is possible that Ferruginous Pygmy-Owls are responding to a decrease in edge habitat within El Salvador, and we recommend further surveys to determine the risk to Ferruginous Pygmy-Owls in response to revegetation management. Poaching may also be a risk for these owls: of eight species of owls confiscated from local markets in 1995 to 2008, the most frequent were Mottled Owls and Ferruginous Pygmy-Owls [@Perez-Leon:2017].

Spectacled Owls followed similar occupancy patterns to that of the Ferruginous Pygmy-Owls, with highest levels of occupancy in the alluvial Nancuchiname Forest. Spectacled Owls can be found in a range of habitats including coffee plantations and urban parks [@Marin-Gomez:2017], and they tend to prefer lowland habitats below 1000 m a.s.l. [@Enriquez-Rocha:1993; @Orihuela-Torres:2018].  Spectacled owls were detected in each of the routes located at low to moderate elevation; however, we also detected Spectacled Owls in Montecristo National Park, transect M-2 (1755 m a.s.l.), which is potentially near the maximum of their elevation tolerance range. Spectacled Owls have been reported to prefer mammalian prey over reptilian or arthropods [@Orihuela-Torres:2018] and preferentially hunt near fallen logs where small mammals often seek refuge [@Esclarski:2014], but we were unable to test for micro-habitat occupancy in our present study. Further refinement over Spectacled Owl habitat selection could be explored in the future to better understand their ecology and distribution in the Neotropics, and specifically, El Salvador.

Nancuchiname Forest was the richest in diversity of our three protected areas; however, many of the habitat specialist species were only detected in Montecristo NP (e.g., Whiskered Owl, Great Horned Owl, Fulvous Owl, and c.f. Stygian Owl). Fulvous Owl is one of El Salvador's 18 endemic bird species [@Komar:2002], and it was detected numerous times in the cloud forest of Montecristo NP (M-1). A Fulvous Owl has also been recently photographed in Montecristo NP [@Gonzalez:2017].

Stygian Owl, which often inhabit dense cloud forests [@Enriquez-Rocha:1993], were listed as an owl species expected to reside in El Salvador [@Perez-Leon:2017], but has been previously undetected in the country. On 26 March 2002, after listening to recordings and viewing photographs, the Montecristo NP guards stated that they thought c.f. Stygian owl was present in the park (personal communication). While there was only one c.f. Stygian Owl detected on our survey routes, we did also hear a c.f. Stygian Owl between the survey stations on 25 April XXXXyear. We also heard c.f. Stygian Owl vocalizations twice on M-2 between station points on 20 March 2005 and once between stations in 2007. We did not record any of these vocalizations by c.f. Stygian Owls, and detections that occurred between stations or away from survey routes were not included in our results. So while we are confident that c.f. Stygian Owls are present in Montecristo NP, we believe that further documentation of the species would be beneficial for confirmation of the species in El Salvador.

During 1979-1981, in the western portion of El Imposible NP, West (1988) heard or observed six owl species, including two that were not detected during this study in El Imposible (Barn Owls and Black-and-white Owls). Barn Owls have been detected in a cave of Loma de Paja Mountain, which was opposite this study's El Imposible routes (Vidal Campos Aguirre, personal communication). Vidal Campos Aguirre also stated that Black-and-white Owls were observed in two areas of the Mistepe River Valley in that same timeframe. It is likely, then, that these two owls still occur in El Imposible, though we did not detect them.

Like many owl species, neotropical owls are difficult to accurately census due to their cryptic nature and nocturnal habits. Owls vocalize to communicate with the same species, to threaten or provoke other species, and to delimit territory [@Johnsgard:1988]. Distinguishing individuals by non-invasive means, such as vocal traits, is preferable when species are rare, sensitive to handling, elusive, or when other techniques are unfeasible [@Terry:2005]. Thus, the vocalizations of crepuscular and cryptic animals such as owls are commonly leveraged in field surveys because they can be diagnostic and are often considered more practically attainable than visual identification [@Konig:1999].

Vocalization surveys for nocturnal birds of prey fall generally into two categories. Spontaneous or “passive” surveys involve listening for animal calls with no prior stimulus, whereas provocation or “broadcast” surveys are conducted by listening after a period of playing a recording of animal vocalizations [e.g., @Baumgardt:2019].  Our surveys coupled both approaches, with spontaneous surveys preceeding broadcast surveys. Overall, the probability of detection was fairly low for all species (median values $\le 0.50$, even though we used locally-recorded owl vocalizations for our broadcast surveys when possible. In Guatemala, @Gerhardt:1989 found that a series of Mottled Owl vocalizations obtained from a non-local ornithological laboratory could not generate a response from Guatemalan owls. We found that the probability of detecting most owls increased after we broadcast the vocalizations of Mottled Owls, Pacific Screech-Owls, Crested Owls, and Spectacled Owls. Conversely, owls were less likely to respond after we played Great Horned Owl calls. Owls are sensitive to interactions with other owls [@Enriquez:1997; @Baumgardt:2019] and may change their response to broadcast calls depending on their life stage (e.g., nesting versus courtship) [@Flesch:2007], so we recommend that surveyors carefully consider the intra- and inter-species interactions that may affect detectability after playing specific bird vocalizations.

Globally, an estimated 75% of the nearly 250 species of owls are associated with dense and undisturbed forests [@Mikkola:2012].  @Perez-Leon:2017 stated that within El Salvador, natural ecosystems such as cloud, deciduous, riparian, and pine-oak forests had the most diverse owl populations, and our findings supported the importance of maintaining protected areas of El Salvador. We found that over the 10-year period of our study, owl species richness and single-species occupancy for three owls were all generally stable in the protected natural areas we investigated. Our study demonstrates the importance of maintaining natural forested areas and diverse habitat heterogeneity in a country under immense climate and population pressures.

@@

![2XX. Probability of occupancy ($\psi_{t,h}$) summarized by route for Ferruginous Pygmy-Owl (FEPO), Mottled Owl (MOOW), and Spectacled Owl (SPEO). Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador: El Imposible National Park (EI-1 and EI-2), Montecristo National Park (M-1 and M-2) and Nancuchiname Forest (N-1 and N-2).  Posteriors were summarized with median $\pm$ 90% credible intervals. \label{psi_means}](../output/figures/psi_means-1.png)

![3XX. Probability of occupancy by route and year for Ferruginous Pygmy-Owl (FEPO), Mottled Owl (MOOW), and Spectacled Owl (SPEO). Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador: El Imposible National Park (EI-1 and EI-2), Montecristo National Park (M-1 and M-2) and Nancuchiname Forest (N-1 and N-2). Posteriors were summarized with median $\pm$ 90% credible intervals. \label{figure2}](../output/figures/psi_byYr-1.png)

@@

![5XX. The probabiliity of detecting owls during spontaneous surveys (i.e., pre-broadcast) and after broadcast, depending on the species of owl used for broadcast vocalization. These logistic regression coefficients were evaluated from the richness model, which combined all detected species (All Species) and the single-species occupancy models for Ferruginous Pygmy-Owls (FEPO), Mottled Owls (MOOW), and Spectacled Owls (SPEO). Owl surveys were conducted from 2003 to 2013 in three different protected areas within El Salvador. Posteriors were summarized with median $\pm$ 90% credible intervals. The pre-broadcase effect size's credible intervals are represented with two grey horizontal lines on each graph \label{figure6}](../output/figures/p_detection-1.png)
